---
title: "**Practical Session PCA<br><font color='#F48225' size='5'>Data visualization (Course 2023/2024)**</font>"
author: "<font color='#A8A8A8' size='3'>Bachelor’s Degree in Bioinformatics</font>"
date: "<font color='#A8A8A8' size='3'>October 2023</font>"
output:
  html_document:
    theme: yeti
    css: https://cdnjs.cloudflare.com/ajax/libs/font-awesome/5.11.2/css/all.css
    self_contained: yes
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 4, fig.align = "center")
```

```{=html}
<style>
  .fig {width: 50%}
  @import url(https://fonts.googleapis.com/css?family=Fira+Sans:300,300i,400,400i,500,500i,700,700i);
  @import url(https://cdn.rawgit.com/tonsky/FiraCode/1.204/distr/fira_code.css);

.center {
  text-align: center;
}

.code_box {
  padding: 1em 1em 1em 1em;
  background: white;
  color: #222222;
  border: 2px solid #222222;
  border-radius: 10px;
  margin-bottom: 10px;
}

.alternative_box {
  padding: 1em;
  background: white;
  color: red4;
  border: 2px solid red;
  border-radius: 10px;
  margin-bottom: 10px;
}

.question_box {
  padding: 1em;
  background: white;
  color: red4;
  border: 2px solid blue;
  border-radius: 10px;
  margin-bottom: 10px;
}

</style>
```
Please, complete all the code proposed and answer to questions. You will
find these on the slides as (\#) and (Q).

You are not expected to include your answer to the check questions (C)
or (K) knowledge thoughts. These are only to guide your progress.

Change the code appropriately or add new lines if necessary.

Provide code alternatives if any.

Add additional interactive plots or additional plots to better interpret
results and generate conclusions.

There are several blocks of code (n=5) to create and several questions
to answer (n=12). Take into account that also a 20% of your grade will
depend on other things: mistakes, clean and neat code, easy to follow,
new elements not covered in this particular class, etc. Check grading
template for details,

::: {.code_box data-latex=""}
**CODE #1: Load, adapt the data and create metadata [2.5 points]**

Load the 2 different replicates provided as separate files into separate
data frames. Check data is correctly loaded and all samples have the
same dimensions. Each file corresponds to a single sample with many
genes (rows) and individual cells (columns).

Column names are single cells ID names corresponding each to a the same
sample.

Load the meta data file (*treatment.txt*) containing information
regarding type of samples for each replicate. The order of the
*treatment.txt* file corresponds to the original order in the
*P1.tsv.gz* and *P2.tsv.gz* files.

Remember to adapt the data accordingly. We need columns of the tables to
represent variables (genes) and rows to be cells. Also, we need to merge
all tables of observations.

What are column names in each dataframe?

Provide code alternatives if any.
:::

### Load data into variables:

First of all, you can use applications external to R to inspect what
type of data you are going to be using and how to deal with it.

Once you have inspected and decided, go for it!

```{r message=FALSE, cache=TRUE}
# --------------------------
## [#1] Read the data
# --------------------------
## Read data into separate variables
P1_data <- read.table(gzfile("./data/P1.tsv.gz"), header = T)
P2_data <- read.table(gzfile("./data/P2.tsv.gz"), header = T)
```


### Check the data

Once loaded into R, check if the data matches what you expected and take
a brief view of the data.

```{r results, include=TRUE, cache=TRUE}
# just print a small head of the data
head(P1_data, n=2)

head(P2_data, n=2)
```

::: {.question_box data-latex=""}
**Q1: What are column names for each original dataframe? Are there
duplicates?**
:::

::: {.question_box data-latex=""}
ANSWER Q1: 
```{r Ex1, echo=F}
cat("Column names of P1:\n")
colnames(P1_data)

cat("Column names of P2:\n")
colnames(P2_data)

cat("Are there any duplicates?\n")
any(duplicated(colnames(P1_data)))
any(duplicated(colnames(P2_data)))
```

:::

### Adapt the data accordingly and merge both tables of observations

Since we need columns of the tables to represent variables and rows to
be samples, individuals, we need to **transpose** the data of the
matrices. Then, we merged experiments into a single data.frame.

```{r, cache=TRUE}
# --------------------------
## [#2] Adapt the data
# --------------------------

#Assign features(genes) to rownames, so that when transposed they are column names and remove the column that contains gene names
##colnames(X1) <- P1_data[,1]
##X1<-X1[-c(1),]

##colnames(X2) <- P2_data[,1]
##X2<-X2[-c(1),]

rownames(P1_data)<-P1_data$gene
P1_data$gene<-NULL

rownames(P2_data)<-P2_data$gene
P2_data$gene<-NULL

#Transposing
X1 <- t(P1_data)
X2 <- t(P2_data)

## Merge
X <- rbind(X1, X2)
```

Now, check that we transformed the data correctly:

```{r results_transform, include=TRUE, cache=TRUE}
dim(X)
```

### Get metadata associated:

Create a dataframe with metadata. We might like to include the names of
the samples as rows, the replicates IDs (1,2 or A,B,...) and the type of
sample as described in the treatment file: T: treated, U: untreated, N:
non-treated.

```{r get_data, eval=TRUE}
# --------------------------
## [#3] Load the metadata
# --------------------------
## Read metadata file and create a factor
treat_file <- as.vector(t(read.table("./data/treatment.txt")))

## create dataframe: add rownames, experiment A & B, and treatment for each cell

## HINT:
## create a vector with 96 treatment types
treatment<-rep(treat_file, 2)

## HINT:
## create a vector with 96 As and 96 Bs
exp_label<-rep(c("A","B"), each=96)

metadata<-cbind(treatment, exp_label)

rownames(metadata)<-rownames(X)
```

::: {.code_box data-latex=""}
**CODE #2: Create a PCA - NOT SCALED: 2.5 points]**

Create PCA and check plots generated to check results:

Perform a PCA on your merged data and plot the results according to the
treatment received, the replicate ID, etc.

*IMPORTANT*: USE `scale=FALSE`.

Use `stats::prcomp` function and remember the PC values are stored
within the 'x' object.

Use several packages (`ggfortify`, base `plot`, `factoextra`) and color
according to treatment and/or experiment.

Provide code alternatives if any.
:::

```{r pca_gen.1, cache=TRUE}
# ------------------------
## [#4] Create PCA results and plot results
# ------------------------

library(ggplot2)
library(ggfortify)
library(factoextra)

#rm("P1_data", "P2_data", "X1", "X2", "exp_label", "treat_file", "treatment")

metadata<-data.frame(metadata)
```


```{r pca_gen.2, eval=FALSE}
saveRDS(X, "X_data.rds")
saveRDS(metadata, "metadata.rds")
```


```{r pca_gen.3, cache=FALSE}
X <- readRDS("X_data.rds")
metadata <- readRDS("metadata.rds")
```


```{r pca_gen.4, cache=TRUE}
pca <- stats::prcomp(X, scale=F)
show(pca)
names(pca)


dim(pca$x)
head(pca$x)

dim(pca$rotation)
head(pca$rotation)
```

Plots
```{r pca_gen.5, cache=TRUE}
# Color by treatment and replicate ID. Use several packages
plot(pca$x, pch=19)
autoplot(pca, metadata, colour = "treatment", shape = "exp_label")
fviz_pca_ind(pca, geom.ind = "point", habillage = metadata$exp_label, addEllipses = TRUE)

ggplot(pca$x, aes(x=PC1, y=PC2, colour=metadata$treatment, shape=metadata$exp_label)) + geom_point()
```



::: {.question_box data-latex=""}
**Q2. According to the PCA plot, explain how samples are distributed.
Justify whether you think that there is any batch effect in our samples
or not associated to the replicate ID.**
:::

::: {.question_box data-latex=""}
**ANSWER Q2**
```{r Ex2, echo=F}

```


:::

### Plot variance explained by each principal component

```{r var_explained, cache=TRUE}
# ------------------------
## [#5] Plot variance for each PC
# ------------------------
# Use several packages

plot()
fviz_pca_ind(pca, geom.ind = "point", habillage = metadata$exp_label, addEllipses = TRUE)

#ns si puc
screeplot(pca)
```

::: {.code_box data-latex=""}
**CODE #3: Create a PCA - SCALED [2.5 points]**

Create PCA and check plots generated to check results:

Perform a PCA on your merged data and plot the results according to the
treatment received, the replicate ID, etc.

*IMPORTANT*: USE `scale=TRUE`.

Use `stats::prcomp` function and remember the PC values are stored
within the 'x' object.

Use several packages (`ggfortify`, base `plot`, `factoextra`) and color
according to treatment and/or experiment.

Provide code alternatives if any.
:::

### Create scaled PCA

```{r scaled_pca, message=TRUE, error=TRUE, cache=TRUE}
# ------------------------
## [#6] Create a scaled PCA
# ------------------------


```

::: {.question_box data-latex=""}
**Q3: Did it work? Why? What does this error mean?**
:::

::: {.question_box data-latex=""}
**ANSWER Q3** 
...
:::

### Fix data and create scaled PCA

Remove those genes that have not been expressed at all in our samples
and try again.

Plot using the scaled dataset the plots using several packages and
treatment and experiment as colouring variables.

```{r fix_scaled_pca, cache=TRUE}
# ------------------------
## [#7] Fix and create a scaled PCA
# ------------------------

## HINT: Remove zero columns (genes not expressed in a cell!)

```

::: {.question_box data-latex=""}
**Q4. Has the batch effect observed been corrected?**\
**If not, is there a clear distinction between batches in any of the
axes (PC1 and PC2)?**
:::

::: {.question_box data-latex=""}
**ANSWER Q4** 
...
:::

::: {.code_box data-latex=""}
**CODE #4: Understanding the batch effect [2.5 points]**

So, the question, "What does it mean for a sample to have a positive
value for this PC?"

We need to find a variable that highly correlates with PC1. We would
check for the total gene expression per cell. In RNA-seq experiments the
data is normalized using different factors such as library size (number
of genes) or gene length.
:::

```{r total_gene, cache=TRUE}
# ------------------------
## [#7] Check the total expression per sample:
# ------------------------

## HINT: The summatory of the rows adds the total expression


```

::: {.question_box data-latex=""}
**Q5. Is there any difference between experiments? Is there any
difference inside the same treated/untreated group? Do you expect cells
with a high value of total expression to have a positive or negative
PC1?**
:::

::: {.question_box data-latex=""}
**ANSWER Q5**

...
:::

#### Check if there are significant correlations

We need to test for significance.

```{r get_correlations, cache=TRUE}
# ------------------------
## [#9] Create a linear regression analysis
# ------------------------


```

::: {.question_box data-latex=""}
**Q6. Are PC1 and total gene expression significantly correlated?** **An
important number of samples seem to have very low or no expression. What
do you think this is due to?**
:::

::: {.question_box data-latex=""}
**ANSWER Q6**
...
:::

::: {.code_box data-latex=""}
**CODE #5: Normalized data accordingly: [2 points]**

After this initial exploration we have realized that we should normalize
each cell by the total expression. We can also use `scale=TRUE` in the
`stats::prcomp` function call.

In RNA-seq experiments the data is normalized using different factors
such as library size (number of genes) or gene length. We could think of
total expression per sample as the factor to use to normalize the data.
:::

```{r norm_data, cache=TRUE}


```

```{r, cache=TRUE}
# ------------------------
## [#10] Create a final representation with the normalized PCA results,
##       using:
##       - color for total expression: Green - Low; Red - High
##       - shape for experiment 
##       - add some additional information: treatment
# ------------------------


```

::: {.question_box data-latex=""}
**Q7. According to the resulting plots, explain what groups can be
distinguished and what the main variables represented by the PC1 and PC2
are and the variance explain by each.**
:::

::: {.question_box data-latex=""}
**ANSWER Q7**
...

:::
